---
title: "PA1_Template.Rmd"
author: "Victoria N"
date: "10 November 2015"
output: html_document
---

# Reproducible Research Peer Assessment 1

Read in the dataset and/or process the data
```{r}
setwd("C:/Users/Vicky/ReproducibleResearch/")
activityData <- read.csv("activity.csv")
```

##What is mean total number of steps taken per day?

Given below is a histogram of the total number of steps taken each day.
```{r}
stepsPerDay <-  tapply(activityData$steps, activityData$date, FUN = sum, na.rm = TRUE)
hist(stepsPerDay, breaks=20)
```

The mean and median of the total number of steps taken per day are:
```{r}
mean(stepsPerDay)
median(stepsPerDay)
```

## What is the average daily activity pattern?
Given below is a time series plot of the average number of steps taken (averaged across all days) versus the 5-minute intervals
```{r}
meanStepsPerInterval <- aggregate(steps ~ interval, data = activityData, FUN= "mean")
plot(meanStepsPerInterval$interval,meanStepsPerInterval$steps,ylab="Average Steps", xlab="Interval", type="l")
```

The 5-minute interval, on average across all the days in the dataset that contains the maximum number of steps is:
```{r}
meanStepsPerInterval[which.max(meanStepsPerInterval$steps),]$interval
```

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

The total number of missing values in the dataset (i.e. the total number of rows with NAs) is:
```{r}
sum(is.na(activityData$steps))
```

The strategy for filling in all of the missing values in the dataset was to replace all of the missing values with the mean for that 5-minute interval. This has been done using an impute.mean function that replaces the na values with the mean value as is shown below.
```{r}
library(plyr)
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
activityData2 <- ddply(activityData, ~ interval, transform, steps = impute.mean(steps))

activityData2 <- activityData2[order(activityData2$date), ]
```

A histogram of the total number of steps taken each day and the mean and median total number of steps taken per day after the missing data has been replaced have been calculated and reported below.
```{r}
stepsPerDay2 <- aggregate(steps ~ date, 
                         data = activityData2, FUN = "sum")
hist(stepsPerDay2$steps, breaks=20)
mean(stepsPerDay2$steps)
median(stepsPerDay2$steps)
```

The mean and median calculated using the data where the missing data has been replaced with the mean for that time interval, are higher than where the missing data was ignored. This is because the missing data would have been treated as zeroes whereas subbing in the mean increases the overall average values.


## Are there differences in activity patterns between weekdays and weekends?

A new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day was created. A panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis) is given below. 
```{r}
dayType <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")) 
        return("weekday") 
    else if (day %in% c("Saturday", "Sunday")) 
        return("weekend") 
}
activityData2$wDay <- sapply(as.Date(activityData2$date), FUN = dayType)
library(ggplot2)
averageStepsByDay <- aggregate(steps ~ interval + wDay, data = activityData2, mean)
ggplot(averageStepsByDay, aes(interval, steps)) + geom_line() + facet_grid(wDay ~ .) + 
    xlab("5-minute interval") + ylab("Number of steps")
```

